Dominance Regions for Rank Two Cluster Algebras

### 2022-02-21 dominance_regions.tex
Some more intro edits

\newcommand{\cC}{\mathcal{C}}
\newcommand{\cD}{\mathcal{D}}
\newcommand{\cI}{\mathcal{I}}
\newcommand{\cP}{\mathcal{P}}
\newcommand{\cQ}{\mathcal{Q}}

@@

\newcommand{\CC}{\mathbb{C}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\ZZ}{\mathbb{Z}}

@@

\title{Dominance Regions for Imaginary $\bfg$-vectors}

@@

\begin{document}
  \begin{abstract}
    We investigate the shapes of the polytopes defining the dominance order for $\bfg$-vectors in cluster algebras of rank two.
    We arrive at a complete description: 
    \begin{itemize}
      \item cluster monomials are never deformable;
      \item the imaginary dominance polytopes are kites, triangles, or trapezoids which extend far outside the imaginary cone, the triangles appearing along rays corresponding to columns of the associated Cartan matrix.
    \end{itemize}
    We conjecture how our observations might generalize to higher dimensions.
  \end{abstract}
  \maketitle

@@

  \section{Introduction}
  Cluster algebras can be understood geometrically by gluing a collection of algebraic tori along birational morphisms known as mutations.
  While such a description seems simple at first glance, this structure has been found across much of the range of mathematical objects under current consideration in the literature.
  To give an idea of how fundamental the process of cluster mutation has come to be understood within mathematics, we note the following (far from exhaustive list of) areas to which this notion is connected: representation theory of artin algebras, especially reflection functors and tilting theory \cite{???}; representation theory of algebraic groups, especially Coxeter groups, double Bruhat cells, and (dual) canonical bases \cite{???}; hyperbolic geometry of surfaces \cite{???}; mathematical physics, especially quiver gauge theories \cite{???}; knot theory, especially the Alexander polynomial \cite{some papers of Schiffler}; algebraic geometry, especially log Calabi-Yau varieties \cite{GHK, GHKK}.
  To say the least, any advance in the fundamental understanding of cluster algebras is likely to find broad application throughout mathematics.

  The investigations of this paper are inspired by a study of Qin on the deformability of bases which are well-behaved under mutations.
  More precisely, we study the notion of dominance order on the tropical labellings for bases of cluster algebras pioneered by Qin.
  As no explicit calculation is carried out in his works, we aim here to initiate a thorough study of the dominance regions for cluster algebras.

  Our calculations reveal the maximal support possible for a pointed basis which remain pointed under all possible mutations.
  In particular, any candidate basis can be ruled out using these results based upon their support alone.

>>>>

\newcommand{\cA}{\mathcal{A}}
\newcommand{\cC}{\mathcal{C}}
\newcommand{\cD}{\mathcal{D}}
\newcommand{\cI}{\mathcal{I}}
\newcommand{\cP}{\mathcal{P}}
\newcommand{\cQ}{\mathcal{Q}}

@@

\newcommand{\CC}{\mathbb{C}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\ZZ}{\mathbb{Z}}

@@

\title{Dominance Regions for Rank Two Cluster Algebras}

@@

\begin{document}
  \begin{abstract}
    We investigate the shapes of the polytopes defining the dominance order for $\bfg$-vectors in cluster algebras of rank two.
    We arrive at a complete description: 
    \begin{itemize}
      \item cluster monomials are never deformable;
      \item the imaginary dominance polytopes are kites, triangles, or trapezoids which extend far outside the imaginary cone; the triangles appearing along rays corresponding to columns of the associated Cartan matrix.
    \end{itemize}
    We conjecture how our observations might generalize to higher dimensions.
  \end{abstract}
  \maketitle

@@

  \section{Introduction}
  Cluster algebras can be understood geometrically by gluing a collection of algebraic tori along birational morphisms known as mutations.
  While such a description seems simple at first glance, this structure has been found across much of the range of mathematical objects under current consideration in the literature.
  To give an idea of how fundamental the process of cluster mutation has come to be understood within mathematics, we note the following (far from exhaustive list of) areas to which this notion is connected: representation theory of artin algebras, especially reflection functors and tilting theory \cite{???}; representation theory of algebraic groups, especially Coxeter groups, double Bruhat cells, and (dual) canonical bases \cite{???}; hyperbolic geometry of Riemann surfaces \cite{???}; mathematical physics, especially quiver gauge theories \cite{???}; knot theory, especially the Alexander polynomial \cite{some papers of Schiffler}; algebraic geometry, especially log Calabi-Yau varieties \cite{GHK, GHKK}.
  To say the least, any advance in the fundamental understanding of cluster algebras is likely to find broad application throughout mathematics.

  The investigations of this paper are inspired by a study of Qin on the deformability of bases which are well-behaved under mutations.
  More precisely, we study the notion of dominance order on the tropical labellings for bases of cluster algebras pioneered by Qin.
  As no explicit calculation is carried out in his works, we aim here to initiate a thorough study of the dominance regions for cluster algebras.
  We begin in the setting of cluster algebras of rank 2, i.e. containing a pair of mutable cluster variables, with no frozen variables.

  Fix integers $b,c>0$ and assume $bc\ge4$.
  When $bc\le3$, the cluster algebra will be of finite-type and the resulting basis of cluster monomials is not deformable \cite{???}.
  The cluster algebra $\cA(b,c)$ is the $\ZZ$-subalgebra of $\QQ(x_1,x_2)$ generated by \emph{cluster variables} $x_m$, $m\in\ZZ$, defined recursively by
  \begin{equation}
    x_{m-1}x_{m+1}=\begin{cases} x_m^b+1 & \text{if $m$ is odd;}\\ x_m^c+1 & \text{if $m$ is even.} \end{cases}
  \end{equation}
  By the Laurent Phenomenon \cite{fomin-zelevinsky}, each $x_m$ is actually an element of $\ZZ[x_k^{\pm1},x_{k+1}^{\pm1}]$ for any $k\in\ZZ$.
  Moreover, these Laurent polynomials are known to have positive coefficients \cite{lee-schiffler,GHKK}.

  Our goal in this note is to explicitly compute the possible bases which are \emph{pointed} with respect to the embedding $\cA(b,c)\into\ZZ[x_k^{\pm1},x_{k+1}^{\pm1}]$ for every $k\in\ZZ$.
  That is, each basis element can be written in the form $\sum\limits_{a_1,a_2 \ge 0} c_{a_1,a_2} x_1^{\lambda_1-ba_1} x_2^{\lambda_2+ca_2}$ with $c_{0,0}=1$ for some $\lambda\in\ZZ^2$ and this structure reproduces after expanding in terms of any pair $\{x_k,x_{k+1}\}$.

  Our calculations reveal the maximal support possible for a pointed basis which remains pointed under all possible mutations.
  In particular, any candidate basis can be ruled out using these results based upon their support alone.

  \begin{theorem}
    The dominance polygon pointed at $\lambda=(\lambda_1,\lambda_2)$ is the region defined by $x\leq \lambda_1, y\geq\lambda_2$, and the following inequalities:
    {
      \everymath={\displaystyle}
      \def\arraystretch{2.8}
      \[
        \begin{array}{rcccl}
          0 & \leq & \frac{b c-\sqrt{b c (b c-4)}}{2 b}(x-\lambda_1)+(y-\lambda_2) & \leq & \max\left(-c\lambda_1-\frac{b c+\sqrt{b c (b c-4)}}{2b}b\lambda_2,0\right)
          \\
          \min\left(-c\lambda_1,0\right) & \leq & \frac{b c-\sqrt{b c (b c-4)}}{2 b}(x-\lambda_1)-(y-\lambda_2) & \leq & 0
          \\
          0 & \leq &  -(x-\lambda_1)-\frac{b c-\sqrt{b c (b c-4)}}{2 c}(y-\lambda_2) & \leq & \max\left(\frac{b c+\sqrt{b c (b c-4)}}{2c}c\lambda_1+b\lambda_2,0\right)
          \\
          \min\left(b \lambda_2,0\right) & \leq & (x-\lambda_1) - \frac{b c-\sqrt{b c (b c-4)}}{2 c} (y-\lambda_2) & \leq & 0
        \end{array}
      \]
    }
  \end{theorem}
